Analyzing Homelessness and Shelter Use in Toronto: A Data-Driven Approach

### 2024-09-26 paper/paper.qmd
Fixes for citation

The dataset is published by the City of Toronto and features daily shelter occupancy and capacity information across various shelter types, such as room-based and bed-based programs. Data collection occurs at 4 a.m. daily, and the dataset includes information on individual shelters' service user counts, the total number of beds or rooms available, and the occupancy rates. The data is refreshed regularly, and the particular dataset used in this paper is up-to-date as of April 23, 2024. Additionally, since this date, the dataset includes occupancy data from hotel-based shelters operated by the Canadian Red Cross Society for refugee claimants. The raw dataset also includes fields such as the shelter group (e.g., adult, family, youth, refugee) and the organization operating the shelter(@torontodailyshelter2).

The data for this paper was sourced from Open Data Toronto and accessed via the opendatatoronto library (@opendatatoronto). The dataset analyzed focuses on daily shelter and overnight service usage in Toronto, providing details on shelter occupancy, capacity, and vacancy (@torontodailyshelter2,). All data analysis was performed in R (@citeR) using a variety of packages, including tidyverse (@thereferencecanbewhatever), here (@here), dplyr (@dplyr), tibble (@tibble), janitor (Sam Firke 2023), ggplot2 (@ggplot2), and knitr (@knitr).

@@

@fig-bed-capacity shed light on the actual program model that exist in these shelters, giving us a deeper look into what resources the large homeless population in Toronto actually needs. The data reveals a significant disparity between Emergency and Transitional shelter capacities. Emergency shelters, designed for short-term relief, saw consistent growth in capacity from 2021 to 2024, reflecting the city’s attempt to address immediate needs. Transitional shelters, which are critical for helping individuals move towards permanent housing, did not experience the same level of capacity expansion. This disparity suggests that while the city is focused on emergency crisis management, it is under-investing in long-term solutions to homelessness(@torontodailyshelter).

With large disparity between the two types, we see a lack of desire in staying in shelter for an extended period of time, which may express the desire to get out of the homeless state or some undesirable nature of these shelters. Recent reports highlight that violence within Toronto’s shelter system has risen significantly, with incidents of interpersonal violence increasing by 283% between 2011 and 2021. This includes physical assaults, verbal abuse, and harassment, which create unsafe environments for both residents and staff. Overcrowding, lack of privacy, and restrictive policies further contribute to the discomfort many shelter users face. These circumstances, combined with the residents' complex mental health and social challenges, may explain why many prefer to leave the shelters promptly. Improving safety and conditions within shelters is essential to making them more welcoming for longer-term stays and providing the necessary support for individuals to move out of homelessness permanently(@cbcsheltersafety).

The gap between the two capacities suggests potential policy directions for the government to focus on, such as increasing support for transitional shelters to provide more long-term solutions. This disparity could also prompt further research into the underlying causes, exploring whether funding limitations, policy focus, or other systemic factors are driving the imbalance.

@@

Shelter use spiked dramatically during the winter months each year, reflecting the severe impact of extreme weather on Toronto's homeless population (@fig-service-user-count). Despite the city's efforts to expand shelter space by adding 400 new spaces and additional warming centers, the system continues to struggle with overwhelming demand. The seasonal fluctuation in service use points to the need for more robust winter preparedness, including additional warming centers, mobile shelters, or other emergency accommodations. By focusing on seasonal fluctuations, the city could better align its homelessness strategy with the predictable increase in demand during the colder months. (@cbctorontohomelesswinter).

In response to the increasing demand for shelters, especially during the winter months, the data in @fig-service-user-count clearly shows significant spikes in shelter usage from November to February each year. These peaks highlight the seasonal pressures on Toronto’s shelter system, with shelter capacity being consistently stretched during colder months. The introduction of additional winter-specific services, such as warming centers and new shelter spaces, may help alleviate some of the strain. However, the persistent upward trend in service user counts over the years underscores that even with these measures, the system continues to face overwhelming demand. The data indicates that without further expansion of both short-term and long-term housing solutions, the shelter system will remain under significant pressure.(@cbctorontowinterhomelessnessplan).

@@

From first glance of @fig-service-sector, we can see the large disparity between men and women organized in the few categories. According to research, by The National Women's Housing & Homelessness Network, based in Canada, average of 36% of homeless population are women. As we compare this data to the graph we see, service used by women do correlate proportionately with their share within the homeless population--while there might be bias within data collected. However, the same research found that 90% of the families using emergency shelters are headed by women--which can lead to a misscalculation in the actual number of women in need of a shelter. We could further theorize that if including women that use shelter within the family unit, the actual disparity between men and women might be smaller that the current number. National Women's Housing & Homelessness Network further suggested that in Canada, there are less shelter dedicated to women than men, men having more that double of dedicated shelters. This is a massive imporvenment that needs to be made in the gender disparity from the policy front, as well as a dataset better assist the imporving rights for homeless women. (@womenshomelessnesscanada)

@@

Additionally, the rising incidents of violence within shelters raise important questions about the safety and effectiveness of these facilities. With reports indicating a 283% increase in interpersonal violence in shelters over the past decade, further research is needed to explore the relationship between shelter conditions and users’ willingness to remain in these spaces(@cbcsheltersafety). Addressing safety concerns, overcrowding, and restrictive policies could make shelters more welcoming and supportive environments.

>>>>

The dataset is published by the City of Toronto and features daily shelter occupancy and capacity information across various shelter types, such as room-based and bed-based programs. Data collection occurs at 4 a.m. daily, and the dataset includes information on individual shelters' service user counts, the total number of beds or rooms available, and the occupancy rates. The data is refreshed regularly, and the particular dataset used in this paper is up-to-date as of April 23, 2024. Additionally, since this date, the dataset includes occupancy data from hotel-based shelters operated by the Canadian Red Cross Society for refugee claimants. The raw dataset also includes fields such as the shelter group (e.g., adult, family, youth, refugee) and the organization operating the shelter[@torontodailyshelter2].

The data for this paper was sourced from Open Data Toronto and accessed via the opendatatoronto library [@opendatatoronto]. The dataset analyzed focuses on daily shelter and overnight service usage in Toronto, providing details on shelter occupancy, capacity, and vacancy [@torontodailyshelter2]. All data analysis was performed in R [@citeR] using a variety of packages, including tidyverse [@thereferencecanbewhatever], here [@here], dplyr [@dplyr], tibble [@tibble], janitor (Sam Firke 2023), ggplot2 [@ggplot2], and knitr [@knitr].

@@

@fig-bed-capacity shed light on the actual program model that exist in these shelters, giving us a deeper look into what resources the large homeless population in Toronto actually needs. The data reveals a significant disparity between Emergency and Transitional shelter capacities. Emergency shelters, designed for short-term relief, saw consistent growth in capacity from 2021 to 2024, reflecting the city’s attempt to address immediate needs. Transitional shelters, which are critical for helping individuals move towards permanent housing, did not experience the same level of capacity expansion. This disparity suggests that while the city is focused on emergency crisis management, it is under-investing in long-term solutions to homelessness[@torontodailyshelter].

With large disparity between the two types, we see a lack of desire in staying in shelter for an extended period of time, which may express the desire to get out of the homeless state or some undesirable nature of these shelters. Recent reports highlight that violence within Toronto’s shelter system has risen significantly, with incidents of interpersonal violence increasing by 283% between 2011 and 2021. This includes physical assaults, verbal abuse, and harassment, which create unsafe environments for both residents and staff. Overcrowding, lack of privacy, and restrictive policies further contribute to the discomfort many shelter users face. These circumstances, combined with the residents' complex mental health and social challenges, may explain why many prefer to leave the shelters promptly. Improving safety and conditions within shelters is essential to making them more welcoming for longer-term stays and providing the necessary support for individuals to move out of homelessness permanently[@cbcsheltersafety].
The gap between the two capacities suggests potential policy directions for the government to focus on, such as increasing support for transitional shelters to provide more long-term solutions. This disparity could also prompt further research into the underlying causes, exploring whether funding limitations, policy focus, or other systemic factors are driving the imbalance.

@@

Shelter use spiked dramatically during the winter months each year, reflecting the severe impact of extreme weather on Toronto's homeless population (@fig-service-user-count). Despite the city's efforts to expand shelter space by adding 400 new spaces and additional warming centers, the system continues to struggle with overwhelming demand. The seasonal fluctuation in service use points to the need for more robust winter preparedness, including additional warming centers, mobile shelters, or other emergency accommodations. By focusing on seasonal fluctuations, the city could better align its homelessness strategy with the predictable increase in demand during the colder months. [@cbctorontohomelesswinter].

In response to the increasing demand for shelters, especially during the winter months, the data in @fig-service-user-count clearly shows significant spikes in shelter usage from November to February each year. These peaks highlight the seasonal pressures on Toronto’s shelter system, with shelter capacity being consistently stretched during colder months. The introduction of additional winter-specific services, such as warming centers and new shelter spaces, may help alleviate some of the strain. However, the persistent upward trend in service user counts over the years underscores that even with these measures, the system continues to face overwhelming demand. The data indicates that without further expansion of both short-term and long-term housing solutions, the shelter system will remain under significant pressure.[@cbctorontowinterhomelessnessplan].

@@

From first glance of @fig-service-sector, we can see the large disparity between men and women organized in the few categories. According to research, by The National Women's Housing & Homelessness Network, based in Canada, average of 36% of homeless population are women. As we compare this data to the graph we see, service used by women do correlate proportionately with their share within the homeless population--while there might be bias within data collected. However, the same research found that 90% of the families using emergency shelters are headed by women--which can lead to a misscalculation in the actual number of women in need of a shelter. We could further theorize that if including women that use shelter within the family unit, the actual disparity between men and women might be smaller that the current number. National Women's Housing & Homelessness Network further suggested that in Canada, there are less shelter dedicated to women than men, men having more that double of dedicated shelters. This is a massive imporvenment that needs to be made in the gender disparity from the policy front, as well as a dataset better assist the imporving rights for homeless women. [@womenshomelessnesscanada]

@@

Additionally, the rising incidents of violence within shelters raise important questions about the safety and effectiveness of these facilities. With reports indicating a 283% increase in interpersonal violence in shelters over the past decade, further research is needed to explore the relationship between shelter conditions and users’ willingness to remain in these spaces[@cbcsheltersafety]. Addressing safety concerns, overcrowding, and restrictive policies could make shelters more welcoming and supportive environments.
